\section{Conclusion}
This report showed that Rust's metaprogramming is capable of implementing design patterns that are identical to manual implementations.
The Abstract Factory pattern was first implemented using Rust attribute macros.
The macro definition led to a foundation that can be used to implement other design pattern macros.
The foundation uses a model to parse the macro input into using \textit{syn}.
This model is then expanded to the pattern's implementation using \textit{quote}.
The foundation was applied to create a function-like Rust macro to implement the Visitor pattern.
The same foundation can thus be applied to other design patterns.

Both macros have some limitations that future research can explore.
The Abstract Factory implementation uses dynamic dispatch, which has a performance trade-off.
``Abstract return types''\footnote{https://www.ncameron.org/blog/abstract-return-types-aka--impl-trait-/} may be a solution to this problem.
Visitor's implementation requires the client to supply the traversal code in the \textit{helper\_tmpl} option manually.
However, Rust macros can read the file system.
Thus, it might be possible for the macro the read all the files for a module and build a composition graph of all the module's types.
The traversal code can then be automatically written by the macro -- effectively reducing the visitor macro call to one line.
Lastly, the Visitor has no global option to apply \textit{no\_default} or \textit{helper\_tmpl} to all its types but will require repeating the option on every type.
The macro currently uses an outer attribute\footnotemark on each type to set the option.
An inner attribute\footnotemark[\value{footnote}] can be used on the macro for global options.

The \textit{iterpolate\_traits} macro creates code that almost forwards to a sub-method by mapping one type to another, close to the Decorator design pattern.
Future research can investigate how a new macro will need to change from \textit{interpolate\_traits} to implement Decorator.
The same goes for the Proxy, Adapter, and some parts of Mediator design patterns.

A study of automating this process will also be interesting.
This will be an external metaprogram that identifies repeated patterns in a library, across libraries, or on whole repositories.
It will then create a Rust macro for each repeated pattern and change the identified code sections to use the macro instead.

\footnotetext{https://doc.rust-lang.org/reference/attributes.html}