\section{Introduction}
This report will focus on the possibility and foundations needed to implement design patterns using metaprogramming in the Rust language.
Software design is constantly faced with solving the same problems.
The design solutions to these problems became known as design patterns.
However, manually programming a design pattern is a tedious task that can be solved using metaprogramming.

Rust is a new programming language created by Mozilla and was announced the most loved language for the 5th consecutive year, according to the 2020 Developer Survey by StackOverflow\footnote{https://insights.stackoverflow.com/survey/2020\#technology-most-loved-dreaded-and-wanted-languages}.
Rust also has metaprogramming support.
Thus, this report's first objective is to see if Rust's metaprogramming has the capacity to implement design patterns.
If Rust's metaprogramming is capable enough, this report will propose a foundation for implementing any design pattern using Rust's metaprogramming.

In \Fref{sec:design-patterns}, this report will explore the need for design patterns and how they came about.
Next, the definition and design of the Abstract Factory pattern and the Visitor pattern will be created.
Both designs will be implemented in Rust in \Fref{sec:reporting}.

\Fref{sec:metaprogramming} will identify different types of metaprogramming and classify the types Rust uses.
Some lesser-used programming concepts present in Rust will be explored in \Fref{sec:rust}.

Finally, \Fref{sec:reporting} will create implementations for the Abstract Factory and Visitor pattern in the same way a programmer will do manually.
The section will then create a Rust macro to write the same Abstract Factory implementation as the manual implementation.
Using the foundations for the Abstract Factory implementation, the section will create a Rust macro for the Visitor implementation.